Virtualisierung eines Echtzeit-Betriebssystems zur Steuerung eines Roboters mit Schwerpunkt auf die Einhaltung der Echtzeit

### 2024-08-18 Masterarbeit_Robotik_Pamuk.tex
Problem and Task Definition++

\begin{itemize}
	\item This work was written at SIGMATEK GmbH \& Co KG, a company that researches, develops, produces, and distributes automation technology for industrial machinery and plant engineering. SIGMATEK uses its own custom Linux-based operating system, to be run on their self-manufactured CPUs. This operating system will be referred to as ``Salamander 4'' in this work. The details of Salamander 4 are explained in Chapter~\ref{cha:salamander4}.
	\item Salamander 4 was created with Yocto~\cite{WelcomeYoctoProject}, an open-source project that helps develop custom Linux-based systems. Yocto is discussed in Section~\ref{sec:yocto}.
	\item Salamander 4 is virtualized through Quick emulator (QEMU), an open-source emulator and virtualizer that allows to run different operating systems on a computer QEMU is described in Section~\ref{sec:qemu}.
	\item Salamander 4 employs hard real-time with Xenomai 3~\cite{XenomaiXenomai}, a real-time framework for Linux that enables deterministic and low-latency performance. Xenomai 3 is detailed in Section~\ref{sec:xenomai}.
	\item The goal is to virtualize Salamander 4 and approach the performance of bare metal through real-time performance tunings, focussing on latency requirements set by SIGMATEK.
	      These modifications are subject of Chapter~\ref{cha:real-time_tuning}.
	\item The latency tool of the Xenomai suite to measure the scheduling latency of the real-time OS was suggested to use by SIGMATEK.
	\item SIGMATEK uses the VARAN bus for communication between machines and systems, a real-time Ethernet network that was specially developed for industrial automation. Its operating principle is outlined in Section \ref{sec:varan}.
	\item For the purpose of replacing the physical CPU, SIGMATEK provided a PCI Insert Card module PCV 522, that was plugged into the PC, as explained in Section \ref{sec:setup_experiment_virtualized}. This component could then be used to communicate with the Input/Output peripherals, just like the physical CPU could when the I/O modules were attached to it. Subsequently, the PCV 522 module of SIGMATEK was connected with the Pulse Width Module PW 022 of SIGMATEK over the VARAN bus and the VI 021 module of SIGMATEK serving as the power supply.
	\item The robotic application from Chapter \ref{cha:robotic_application} was written in Lasal Class 2, an object oriented programming tool of SIGMATEK with client-server technology and graphic representation.
\end{itemize}

@@

\bigskip \noindent In the course of writing this thesis, the white paper "Real-Time Performance Tuning Best Practice Guidelines for KVM-Based Virtual Machines"~\cite{RealTimePerformanceTuning2022} by Intel has been instrumental. Given that the working computer used for this work is equipped with an Intel processor, the guidelines presented in this document were particularly relevant and valuable for optimizing real-time performance in KVM-based virtual machines. Chapter \ref{cha:real-time_tuning} contains an extensive tuning process, which involves configurations spanning the BIOS, kernel, host OS, QEMU/KVM virtualization layer, and the Salamander 4 OS itself. The document concludes that the described tuning methods effectively improve real-time performance for KVM-based VMs, even under heavy workloads.

@@

@@

\section{Problem and Task Definition}

\section{Objective}

@@

\chapter{Theoretical Foundations}
\section{Yocto}\label{sec:yocto}

@@

As a starting point, initial latency values of both the bare metal and virtualization versions were measured with the \texttt{latency} tool of the xenomai tool suite. Salamander 4 bare metal refers to the proprietary hardware of SIGMATEK used to employ the custom operating system. Salamander 4 virtualization refers to a virtual version of the Salamander 4 hardware platform, achieved through QEMU/KVM. Sections~\ref{sec:salamander4-bare-metal} and~\ref{sec:salamander4-virtualization} specify the details of the measurements for both versions. In the further course, the aim was to bring the latency values of the virtualization closer to those of the hardware and guarantee deterministic and reliable behavior.

\section{Salamander 4 Bare Metal}\label{sec:salamander4-bare-metal}
The output of the command~\texttt{uname -a} for Salamander 4 bare metal is shown in Code~\ref{output:uname_a_hw}.

\section{Salamander 4 Virtualization}\label{sec:salamander4-virtualization}
In addition to providing Salamander 4 on its own hardware, SIGMATEK has also developed a virtualised version of this operating system. It was developed using Yocto, an open-source project that allows customised Linux distributions to be created for embedded systems. Upon generating the necessary files, Yocto provides a QEMU folder with the following components shown in Code~\ref{lst:ls}. QEMU is the environment in which the virtualization runs, as it is an open-source tool for hardware virtualization~\cite{QEMU}.

\noindent The inital QEMU script after the custom Yocto build and the starting point for this work is shown in Code~\ref{script:qemu_def}. This script is used to start QEMU with required parameters to boot Salamamder 4 OS. It will be adjusted in Chapter~\ref{cha:real-time_tuning} in order to accompany real-time performance tunings.

\chapter{Real-Time Performance Tuning}\label{cha:real-time_tuning}

In this chapter, the significant initial gap in latency statistics between the virtualized system and the bare metal system is tackled. For this reason, an extensive tuning process is carried out. This involves configurations spanning the BIOS, kernel, host OS, QEMU/KVM virtualization layer, and the Salamander 4 OS itself. The individual configurations will be discussed in detail and the modifications will be justified with a clear explanation. The goal is to bring the latency of the virtualized system closer to that of the bare metal system, thereby ensuring deterministic behavior under real-time constraints.

@@

\bigskip \noindent Nevertheless, a real-time kernel alone does not make a system truly “real-time”~\cite{WhatRealtimeLinux}. Additional modifications are required to achieve this. The next sections will deal with these real-time performance tunings. The very first step in this regard is the isolation of a CPU to dedicate it solely to the desired real-time task. CPU isolation and affinity will be covered in detail later in Section~\ref{subsec:cpu_isolation}, but its impact will be briefly demonstrated here.

@@

After each tuning, the latency test of Xenomai will be executed to evaluate the improved latency of the system.

@@

\subsection{Tune LAPIC Timer Advance}
The Local Advanced Programmable Interrupt Controller (LAPIC) is a built-in timer that handles the delivery of interrupts to the CPU.
It generates interrupts at a rate. This rate can be tuned in the \texttt{/sys/module/kvm/parameters/lapic\_timer\_advance\_ns} file to reduce the frequency of interrupts and therefore decrease the latency of the guest VM. The default value is ``-1'', which means that the kernel will automatically calculate an appropriate advance for the timer. Here, it is set to the value ``7500''. Hence,the timer interrupt will be delivered 7500 nanoseconds earlier than it is actually due. This gives the VM more time to handle it.

@@

~\cite{canbazPerformanceAnalysisRealtime2022}
~\cite{malallahComprehensiveStudyKernel2021}
~\cite{wangRealtimeEmbeddedSystems2017}
~\cite{buttazzoHardRealtimeComputing2024}
~\cite{lipariRealTimeSchedulingHard}
~\cite{perneelRealtimeCapabilitiesStandard2015}
~\cite{reghenzaniRealTimeLinuxKernel2020}
~\cite{adamRealTimePerformanceResponse2021}
~\cite{RealTimePerformanceTuning2022}
~\cite{yoonRealTimePerformanceAnalysis}
~\cite{cinqueVirtualizingMixedCriticalitySystems2022}
~\cite{sandstromVirtualizationTechnologiesEmbedded2013}
~\cite{taccariEmbeddedRealTimeVirtualization2014}
~\cite{javierperezHowRealTime2022}
~\cite{guStateoftheArtSurveyRealTime2012}
~\cite{garcia-vallsChallengesRealtimeVirtualization2014}
~\cite{scordinoRealTimeVirtualizationIndustrial2020}
~\cite{maPerformanceTuningKVMbased}
~\cite{junzhangPerformanceAnalysisKVMBased2010}
~\cite{broskyShieldedProcessorsGuaranteeing2003}
~\cite{kiszkaLinuxRealTimeHypervisor}
~\cite{mckenneyRealTimeVs}
~\cite{deoliveiraDemystifyingRealTimeLinux}
~\cite{huang2015performance}
~\cite{pielAsymmetricSchedulingLoad2006}
~\cite{queirozTestingLimitsGeneralpurpose2023}
~\cite{kirovaImpactModernVirtualization2019}
~\cite{REALTIMEOPERATING}
~\cite{masrurVMBasedRealTimeServices2010}
~\cite{HardRealTime}
~\cite{adamPerformanceAssessmentLinux2021}
~\cite{jiangRealTimeSystemManycore}
\clearpage

>>>>

\begin{itemize}
	\item This work was written at SIGMATEK GmbH \& Co KG, a company that researches, develops, produces, and distributes automation technology for industrial machinery and plant engineering. SIGMATEK uses its own custom Linux-based operating system, to be run on their self-manufactured CPUs. This operating system will be referred to as ``Salamander 4'' in this work. The details of Salamander 4 are explained in Chapter~\ref{cha:salamander4}.
	\item Salamander 4 was created with Yocto~\cite{WelcomeYoctoProject}, an open-source project that helps develop custom Linux-based systems. Yocto is discussed in Section~\ref{sec:yocto}.
	\item Salamander 4 is virtualized through Quick emulator (QEMU), an open-source emulator and virtualizer that allows to run different operating systems on a computer QEMU is described in Section~\ref{sec:qemu}.
	\item Salamander 4 employs hard real-time with Xenomai 3~\cite{XenomaiXenomai}, a real-time framework for Linux that enables deterministic and low-latency performance. Xenomai 3 is detailed in Section~\ref{sec:xenomai}.
	\item The goal is to virtualize Salamander 4 and approach the performance of bare metal through real-time performance tunings, focussing on latency requirements set by SIGMATEK.
	      These modifications are subject of Chapter~\ref{cha:real-time_tunings}.
	\item The latency tool of the Xenomai suite to measure the scheduling latency of the real-time OS was suggested to use by SIGMATEK.
	\item SIGMATEK uses the VARAN bus for communication between machines and systems, a real-time Ethernet network that was specially developed for industrial automation. Its operating principle is outlined in Section \ref{sec:varan}.
	\item For the purpose of replacing the physical CPU, SIGMATEK provided a PCI Insert Card module PCV 522, that was plugged into the PC, as explained in Section \ref{sec:setup_experiment_virtualized}. This component could then be used to communicate with the Input/Output peripherals, just like the physical CPU could when the I/O modules were attached to it. Subsequently, the PCV 522 module of SIGMATEK was connected with the Pulse Width Module PW 022 of SIGMATEK over the VARAN bus and the VI 021 module of SIGMATEK serving as the power supply.
	\item The robotic application from Chapter \ref{cha:robotic_application} was written in Lasal Class 2, an object oriented programming tool of SIGMATEK with client-server technology and graphic representation.
\end{itemize}

@@

@@

\bigskip \noindent In the course of writing this thesis, the white paper "Real-Time Performance Tuning Best Practice Guidelines for KVM-Based Virtual Machines"~\cite{RealTimePerformanceTuning2022} by Intel has been instrumental. Given that the working computer used for this work is equipped with an Intel processor, the guidelines presented in this document were particularly relevant and valuable for optimizing real-time performance in KVM-based virtual machines. Chapter \ref{cha:real-time_tunings} contains an extensive tuning process, which involves configurations spanning the BIOS, kernel, host OS, QEMU/KVM virtualization layer, and the Salamander 4 OS itself. The document concludes that the described tuning methods effectively improve real-time performance for KVM-based VMs, even under heavy workloads.

@@

\clearpage
\section{Problem and Task Definition}
As stated previously, virtualization of real-time systems has many advantages in comparison to traditional hardware-based approaches. Solutions based on pure hardware are often built for a distinct purpose and are hence limited in flexibility. It is difficult to adapt the hardware topology to new requirements or new applications. Moreover, during periods when the demand is lower, expensive hardware may sit idle, leading to inefficient use of resources and a poor return on investment~\cite{gabrielResourceAwareParameterTuning2019}. Another crucial factor is of course the high cost factor associated with procuring and maintaining separate hardware for each real-time system~\cite{bhardwajVirtualizationCloudComputing2021}. Especially when organizations scale their operations, they have to add more physical machines, and the expenses rise rapidly~\cite{abbasiExploringOpenStackScalable2023}. Besides, one needs physical access to the hardware for any updates or maintenance, which is hard to achieve, especially for systems in remote or difficult-to-access locations. This can lead to system downtime and may not only disrupt the operations but can also result in lost productivity and potential revenue~\cite{mahfoudRealTimePredictiveMaintenanceBased2024}.

\bigskip \noindent Virtualization promises to overcome these challenges \cite{queirozContainerbasedVirtualizationRealtime2024} \cite{bhardwajVirtualizationCloudComputing2021a} \cite{cinqueEvaluatingVirtualizationFog2023}. However, virtualization also has its own challenges and pitfalls. The additional layers of abstraction can lead to increased overhead and latency \cite{casiniLatencyAnalysisVirtualization2021} \cite{zhangEvaluatingOptimizingVirtualization2010}. This is not desired and can be particularly problematic for real-time systems that require strict timing constraints. Making sure that tasks run predictably and on time in a virtualized environment, can be challenging. The hypervisor and other virtual components can cause variations in how long tasks take to execute \cite{garcia-vallsChallengesRealtimeVirtualization2014}. When there are multiple virtual machines (VM) sharing the same physical resources, the VMs can affect each other's performance and consequently violate real-time constraints \cite{queirozContainerbasedVirtualizationRealtime2024}.

\bigskip \noindent In this master’s thesis, the task is the virtualization of a real-time operating system to control a robot, with a focus on compliance with real-time determinism. Chapter \ref{cha:methodology} explains in detail the procedure, how this was tackled.  As already established in Section~\ref{sec:application_context}, Salamander 4 is built with Yocto, employs hard real-time with Xenomai 3 and virtualized through QEMU/KVM.  These components are briefly described in Chapter \ref{cha:theoretical_foundations}. The initial latency values are measured with the \texttt{latency} program of the Xenomai tool suite. The initial situation is described in Section \ref{sec:starting_situation}. In a nutshell, there is a significant initial gap in latency statistics between the virtualized Salamander 4 and the Salamander 4 on bare metal. For this reason, an extensive tuning process is carried out to achieve real-time performance and determinism. These modifications involve configurations spanning the BIOS, kernel, host OS, QEMU/KVM virtualization layer, and the Salamander 4 OS (guest) itself. The individual configurations are discussed in detail in Chapter \ref{cha:real-time_tunings} and the modifications are justified with a clear explanation.

\section{Objective}
The goal is to bring the latency of the virtualized Salamander 4 closer to that of the bare metal system, thereby ensuring deterministic behavior under real-time constraints.

\section{Starting Situation}\label{sec:starting_situation}
As a starting point, initial latency values of both the bare metal and virtualization versions were measured with the \texttt{latency} tool of the xenomai tool suite. Salamander 4 bare metal refers to the proprietary hardware of SIGMATEK used to employ the custom operating system. Salamander 4 virtualization refers to a virtual version of the Salamander 4 hardware platform, achieved through QEMU/KVM. Sections~\ref{sec:salamander4-bare-metal} and~\ref{sec:salamander4-virtualization} specify the details of the measurements for both versions. In the further course, the aim was to bring the latency values of the virtualization closer to those of the hardware and guarantee deterministic and reliable behavior.

\subsection{Salamander 4 Bare Metal}\label{sec:salamander4-bare-metal}
The output of the command~\texttt{uname -a} for Salamander 4 bare metal is shown in Code~\ref{output:uname_a_hw}.

\subsection{Salamander 4 Virtualization}\label{sec:salamander4-virtualization}
In addition to providing Salamander 4 on its own hardware, SIGMATEK has also developed a virtualised version of this operating system. It was developed using Yocto, an open-source project that allows customised Linux distributions to be created for embedded systems. Upon generating the necessary files, Yocto provides a QEMU folder with the following components shown in Code~\ref{lst:ls}. QEMU is the environment in which the virtualization runs, as it is an open-source tool for hardware virtualization~\cite{QEMU}.

\noindent The inital QEMU script after the custom Yocto build and the starting point for this work is shown in Code~\ref{script:qemu_def}. This script is used to start QEMU with required parameters to boot Salamamder 4 OS. It will be adjusted in Chapter~\ref{cha:real-time_tunings} in order to accompany real-time performance tunings.

@@

\chapter{Theoretical Foundations}\label{cha:theoretical_foundations}
\section{Yocto}\label{sec:yocto}

@@

\chapter{Real-Time Performance Tuning}\label{cha:real-time_tunings}

In this chapter, the significant initial gap in latency statistics between the virtualized system and the bare metal system is tackled. For this reason, an extensive tuning process is carried out. This involves configurations spanning the BIOS, kernel, host OS, QEMU/KVM virtualization layer, and the Salamander 4 OS itself. The individual configurations are discussed in detail and the modifications are justified with a clear explanation. The goal is to bring the latency of the virtualized system closer to that of the bare metal system, thereby ensuring deterministic behavior under real-time constraints.

@@

\bigskip \noindent Nevertheless, a real-time kernel alone does not make a system truly “real-time”~\cite{WhatRealtimeLinux}. Additional modifications are required to achieve this. The next sections will deal with these real-time performance tunings. The very first step in this regard is the isolation of a CPU to dedicate it solely to the desired real-time task. CPU isolation and affinity are covered in detail later in Section~\ref{subsec:cpu_isolation}, but its impact are briefly demonstrated here.

@@

After each tuning, the latency test of Xenomai is executed to evaluate the improved latency of the system.

@@

\subsection{Tune LAPIC Timer Advance}
The Local Advanced Programmable Interrupt Controller (LAPIC) is a built-in timer that handles the delivery of interrupts to the CPU.
It generates interrupts at a rate. This rate can be tuned in the \texttt{/sys/module/kvm/parameters/lapic\_timer\_advance\_ns} file to reduce the frequency of interrupts and therefore decrease the latency of the guest VM. The default value is ``-1'', which means that the kernel will automatically calculate an appropriate advance for the timer. Here, it is set to the value ``7500''. Hence, the timer interrupt will be delivered 7500 nanoseconds earlier than it is actually due. This gives the VM more time to handle it.

@@

~\cite{canbazPerformanceAnalysisRealtime2022}
~\cite{malallahComprehensiveStudyKernel2021}
~\cite{wangRealtimeEmbeddedSystems2017}
~\cite{buttazzoHardRealtimeComputing2024}
~\cite{lipariRealTimeSchedulingHard}
~\cite{perneelRealtimeCapabilitiesStandard2015}
~\cite{reghenzaniRealTimeLinuxKernel2020}
~\cite{adamRealTimePerformanceResponse2021}
~\cite{RealTimePerformanceTuning2022}
~\cite{yoonRealTimePerformanceAnalysis}
~\cite{cinqueVirtualizingMixedCriticalitySystems2022}
~\cite{sandstromVirtualizationTechnologiesEmbedded2013}
~\cite{taccariEmbeddedRealTimeVirtualization2014}
~\cite{javierperezHowRealTime2022}
~\cite{guStateoftheArtSurveyRealTime2012}
~\cite{garcia-vallsChallengesRealtimeVirtualization2014}
~\cite{scordinoRealTimeVirtualizationIndustrial2020}
~\cite{maPerformanceTuningKVMbased}
~\cite{junzhangPerformanceAnalysisKVMBased2010}
~\cite{broskyShieldedProcessorsGuaranteeing2003}
~\cite{kiszkaLinuxRealTimeHypervisor}
~\cite{mckenneyRealTimeVs}
~\cite{deoliveiraDemystifyingRealTimeLinux}
~\cite{gabrielResourceAwareParameterTuning2019}
~\cite{bhardwajVirtualizationCloudComputing2021}
~\cite{abbasiExploringOpenStackScalable2023}
~\cite{mahfoudRealTimePredictiveMaintenanceBased2024}
~\cite{queirozContainerbasedVirtualizationRealtime2024}
~\cite{bhardwajVirtualizationCloudComputing2021a}
~\cite{cinqueEvaluatingVirtualizationFog2023}
~\cite{huang2015performance}
~\cite{pielAsymmetricSchedulingLoad2006}
~\cite{queirozTestingLimitsGeneralpurpose2023}
~\cite{kirovaImpactModernVirtualization2019}
~\cite{REALTIMEOPERATING}
~\cite{masrurVMBasedRealTimeServices2010}
~\cite{HardRealTime}
~\cite{adamPerformanceAssessmentLinux2021}
~\cite{jiangRealTimeSystemManycore}
\clearpage
